Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-08-04 manuscript.RMD
fixed p-value format in manuscript

---
title: "Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults"
author: "Kristian Lian^1^, Sara Christine Olsen Moen^1^, Håvard Hamarsland^1^, Daniel Hammarström^1^, Stian Ellefsen^1^"
bibliography: resources/ribose-paper.bib
csl: resources/jap.csl
output:
  word_document:
    reference_docx: resources/reference-template-02.docx
editor_options: 
  chunk_output_type: console
---

@@

```{r}
#| echo: false

@@

**Introduction.** Muscle hypertrophy, induced by resistance training (RT), is determined by ribosomal biogenesis. Synthesis of ribosomes may augment the transcription of ribosomal RNA, as evident from in vitro studies. However, its effect on RT-induced adaptations remains unexplored.  
**Methods.** Sixteen healthy, moderately trained individuals (male/female, n = XX/XX; age, mean (sd)),<!-- Add male/female numbers, age--> participated in a within-participant trial with unilateral resistance training performed with glucose- and placebo supplement on alternate days over six sessions per condition. Participants remained fasted overnight, receiving glucose/placebo and protein supplementation only before and after RT sessions (3 sets of 10 repetitions maximum unilateral leg press and knee extension). Micro biopsies were sampled from m. vastus lateralis pre and post the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured before, during (days 4, 5, 8, 9) and after the intervention. 
**Results.** <!-- update statistics in this section -->RT led to robust increases in total RNA and rRNA in glucose and placebo (RNA: 26 and 22% respectively, rRNA: 33-43% and 33-41% respectively), with no differences between supplements (p > 0.05). There were no differences in content of c-Myc (*p* = 0.160), UBF (*p* = 0.154) and rpS6 (*p* = 0.386) between treatments in response to RT. A linear relationship was found between content of UBF and total RNA, where an increase of 1 SD unit in UBF equated to 6.8% increase in total RNA (*p* = 0.002). Lastly, there were no differences between supplements in maximal unilateral isometric knee extension force and isokinetic peak torque during the intervention (p > 0.05) and following the last session (p > 0.05), apart from the 60 d/s test before last RT session (p < 0.05). 
**Conclusion.** Glucose supplement before and after RT does not augment RT-induced ribosome biogenesis, nor does it enhance muscular performance and recovery after two weeks of heavy RT in moderately trained, young individuals.

@@

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition affecting the internal physiological milieu and subsequently affecting the ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Check these refs, do they correspond to genet epigen and internal mileu? Kristian: The way I read it, it looks like it corresponds --> Although RT program modifications such as altered exercise volume or intensity may induce more favourable adaptations in some individuals, these benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017].<!-- Add ref for more rt program modifications. Kristian: Added krieger et al. 2009 and schoenfeld et al. 2017 --> For instance, although changes in training volume generally induce greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seem necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? Im not sure, are you thinking about "other means than modification of RT" being contradictory of "modifications such as altered exercise volume.." sentence? I think there is a point to establishing that yes, increasing exercise volume is a potent way to increase the RT stimulus in general. However, in cases where this alone fails to elicit true responses, other means are necessary. Or am I missinterpreting this? -->

Nutritional supplements such as protein and creatine effectively optimize RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in the*de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] <!--Add ref that indicate high energy requirement for ribo biog possibly (VONWALDEN19;TANAKA18? MOSS. Kristian: These three should do it-->which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by Pol I is considered the limiting step in synthesizing new ribosomes [@moss_promotion_1995].  Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target S6K1 serve as a mediator from insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of the translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promotes rRNA transcription in cell cultures[@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

@@

The study was designed as a 12-day double-blinded randomized controlled trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with glucose and one leg with a placebo (Figure 1A).
<!--from the "old" methods: either their dominant or non-dominant leg with glucose or placebo before and after RT, and whether they started RT with glucose or placebo. Thus, participants exercised every day, alternating between e.g., one leg + glucose and the other leg + placebo - this might be TMI, as it's presented in the figure as well?--> One person was exclusively responsible for the randomization code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucosum monohydricum vs. ~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardize this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose/placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between treatments. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of PRO supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously.  During sessions, participants were free to ingest water ad libitum.

@@

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at pre-intervention testing during familiarization (Figure 1, Ddays -7 and -5). The participants performed a general warm-up with 10min cycling on an indoor exercise bicycle. A protocol consisting of 1x10, 1x6 and 1x3 repetitions with a load equivalent to ~50-75% of assumed max repetitions, was used as a specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, with a knee angle of 45 degrees (start), within a 2.5kg range. Attempts, where participants did not reach 45 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up and pause protocol and was defined as maximal load lifted in a controlled fashion, reaching full extension of the knee joint, within a 1.25kg range. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minutes of rest were given during the specific warm-up, and three minutes of rest was given between 1RM attempts.

### Resistance training protocoL

@@

To measure blood glucose levels with and without glucose ingestion/training, capillary blood was collected from finger sticks on days with biopsy sampling. One capillary blood sample was taken on day 1 (Figure 1A) to serve as a baseline. On days 11 and 12 (Figure 1A), capillary blood samples were collected seven times: I) Immediately prior to protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5hrs after protein ingestion (0800hrs, i.e., immediately before GLU/PLAC intake), IV) 2hrs after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (~0915hrs), V) immediately after training (~0930hrs), and VI) 2hrs after completion of training (~1130hrs). Capillary blood samples were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the capillary samples except 45min after protein ingestion and in the middle of the RT session, to analyse endocrine variables.

@@

After 5 RT sessions (Figure 1D, Post 5RT), there was a significant difference in strength between glucose and placebo,  where placebo decreased strength by 7% more than glucose (Figure 1D, *p* = `r pdiff.c4`). At the other time points, there were no significant differences in mean change between exercising with glucose and placebo (Figure 1D, Post 2RT: *p* = `r pdiff.c2`, Post 4RT: *p* = `r pdiff.c3`, 30min post 6RT: *p* = `r pdiff.c5`, 2h post 6RT: *p* = `r pdiff.c6`, 23h post 6RT: *p* = `r pdiff.c7`). Both RT with glucose and placebo led to significantly reduced strength post-fifth session compared to baseline, by 11 and 18% respectively (Figure 1D, Post 5RT: *p* = 0.000). Comparisons of the acute data gathered from the post-fifth session until and including 23hrs post-sixth session showed a significant mean increase in strength of 5-9% after RT with glucose and placebo 30min after the sixth RT session (Figure 1D, 30min post 6RT: *p* = `r apc2`) and two hours after the sixth RT session (Figure 1D, 2h post 6RT: *p* = `r apc3`). Twenty-three hours after the last sixth RT session, strength was unchanged compared to after the fifth RT session (Figure 1D, 23h post 6RT: *p* = `r apc4`).

@@

@@

@@

stat_west <- lapply(prot_models$prot_models, west_estimate)
p_west <- unlist(lapply(prot_models$prot_models, west_estimate, p = TRUE))

@@

corr_incr <- round(100 * (exp(coef(summary(mod))[2,1])-1),1)
corrp <- sprintf("%.3f", coef(summary(mod))[2,5])

@@

corrcoef <- coef(summary(corr.mod))
corrp <- round(corrcoef[2,5], 3)

@@

The main findings of the present study were that heavy resistance training with glucose did not affect markers of ribosome biogenesis, measured as total RNA, rRNA, and rDNA transcription initiation-associated proteins, compared to RT with placebo. Similarly, RT with glucose in general did not affect markers of skeletal muscle functionality such as strength and recovery, or total training session volume. At one time point, RT with glucose led to less reduction in strength compared to placebo. Levels of plasma glucose and c-peptide were significantly higher before, during and after the RT sessions with glucose. Importantly, there were no differences in daily macronutrient intake on pairwise consecutive days. These findings suggest that the effects of high vs. low glucose/glucose starvation treatment on rDNA transcription initiation observed in previous *in vitro* studies [@mariappan_ribosomal_2011; @tanaka_mild_2015] are not translatable to acute effects in human skeletal muscle *in vivo*. Notably, the present study was specifically designed to investigate differences in pre-post changes within participants, and is as such also quite limited to this perspective. Hence, the observed effects of time serve as supplementary to the efficiency of the design, without a negative control group to compare to.

>>>>

---
title: "Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults"
author: "Kristian Lian^1^, Sara Christine Olsen Moen^1^, Håvard Hamarsland^1^, Daniel Hammarström^1^, Stian Ellefsen^1^"
bibliography: resources/ribose-paper.bib
csl: resources/jap.csl
output:
  word_document:
    reference_docx: resources/jap-template.docx
editor_options: 
  chunk_output_type: console
---

@@

<!-- To do: Fix tables, add outlier figures to the readme -->
```{r}
#| echo: false

@@

```{r abstract in-line data}

## total_rna_stats
total_rna_stats <- readRDS("./data/data-gen/rna/total_rna_stats.RDS")

## stat_rrna
# stat_rrna$trg_47s, stat_rrna$trg_18s, stat_rrna$trg_28s, stat_rrna$trg_5.8s, stat_rrna$trg_5s
stat_rrna <- readRDS("./data/data-gen/rna/stat_rrna.RDS")

## stat_west
#stat_west$cmyc, stat_west$ubf, stat_west$rps6
stat_west <- readRDS("./data/data-gen/protein/stat_west.RDS")

## p_west
# min(p_west) - max(p_west)
p_west <- readRDS("./data/data-gen/protein/p_west.RDS")

## corr_incr
# $\sim$ `r corr_incr`
corr_incr <- readRDS("./data/data-gen/protein/corr_incr.RDS")
corrp <- readRDS("./data/data-gen/protein/corrp.RDS")

**Introduction.** Muscle hypertrophy, induced by resistance training (RT), is determined by ribosomal biogenesis. Synthesis of ribosomes may augment the transcription of ribosomal RNA, as evident from in vitro studies. However, its effect on RT-induced adaptations remains unexplored.

**Methods.** Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 23.7 (1.8)/24.6 (4.8)),<!-- Add male/female numbers, age. Kristian: Done--> participated in a within-participant trial with unilateral resistance training performed with glucose- and placebo supplement on alternate days over six sessions per condition. Participants remained fasted overnight, receiving glucose/placebo and protein supplementation only before and after RT sessions (3 sets of 10 repetitions maximum unilateral leg press and knee extension). Micro biopsies were sampled from m. vastus lateralis pre and post the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured before, during (days 4, 5, 8, 9) and after the intervention.

**Results.** <!-- update statistics in this section -->From baseline to post-intervention total RNA and rRNA per unit muscle weight increased on average by ~24% and ~33-43%, respectively (Figure 2), with no differences between supplements (Figure 2A; mean difference `r total_rna_stats`, Figure 2B; 47S, `r stat_rrna$trg_47s`; 18S, `r stat_rrna$trg_18s`; 28S, `r stat_rrna$trg_28s`; 5.8S, `r stat_rrna$trg_5.8s`; 5S, `r stat_rrna$trg_5s`). RT with glucose resulted in estimated levels of c-Myc, UBF and RPS6 being `r stat_west$cmyc`, `r stat_west$ubf` and `r stat_west$rps6`% lower compared to placebo, respectively, without showing statistical significance (*p* = `r min(p_west)`-`r max(p_west)`; Figure 3A). There was a linear relationship between UBF and total RNA, where an increase of 1 SD unit of UBF equated to a $\sim$ `r corr_incr`% increase in total RNA (*p* = `r corrp`; Figure 3C). Lastly, there were no differences between supplements in maximal unilateral isometric knee extension force and isokinetic peak torque during the intervention (p > 0.05) and following the last session (p > 0.05), apart from the 60 d/s test before last RT session (p < 0.05).

**Conclusion.** Glucose supplement before and after RT does not augment RT-induced ribosome biogenesis, nor does it enhance muscular performance and recovery after two weeks of heavy RT in moderately trained, young individuals.

<!--*Introduction:*Responses to resistance training (RT) are not uniform across the population, and individuals responding poorly to RT show a blunted ability to produce novel ribosomes [@hammarstrom_benefits_2020; @stec_ribosome_2016]. Recent research solidifies the relationship between increased rates of total RNA and the magnitude of RT-induced adaptations [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_revisiting_2019], with ribosome abundance being associated with UBF protein levels [@hammarstrom_ribosome_2022]. However, 10-15% of individuals undertaking a period of RT showed considerable impaired growth [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Since modifying RT alone does not convert low-responders to high-responders [@hammarstrom_benefits_2020] other means than RT *per se* is necessary to circumvent this discrepancy. High glucose treatment has been observed to augment rDNA transcription in vitro [@sebata_control_2018], however, its effect on RT-induced adaptations remains quite unexplored.  
*Methods:*Sixteen healthy moderately trained participants were included and randomized to unilateral resistance training with glucose- and placebo supplement. Participants remained fasted overnight receiving glucose/placebo and protein only before and after RT. Resistance training consisted of 3 sets of 10 repetitions maximum unilateral leg press and knee extension. Micro biopsies were sampled pre (T1/T2) and post (T3/T4) the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured pre (T0), during (days 4, 5, 8, 9) and post (T3/T4, day 13) the intervention.  
*Results:*Despite robust increases in total RNA and rRNA in glucose and placebo (RNA: 26 and 22% respectively, rRNA: 33-43% and 33-41% respectively), there were no differences between supplements (p > 0.05). There were no differences in change of c-Myc, UBF and rpS6 content between treatments (*p* > 0.05). There was indeed a linear relationship between content of UBF and total RNA, where an increase of 1 SD unit in UBF equated to 6.8% increase in total RNA (*p* = 0.002). Lastly, there were no differences between supplements in maximal unilateral isometric knee extension force and isokinetic peak torque during the intervention (p > 0.05) and following the last session (p > 0.05), apart from the 60 d/s test before last RT session (p < 0.05). 
*Conclusion:* Glucose supplement before and after RT does not acutely augment ribosome biogenesis, nor does it enhance muscular performance and recovery after two weeks of heavy RT in moderately trained young individuals. OLD ABSTRACT-->

@@

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition affecting the internal physiological milieu and subsequently affecting the ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Check these refs, do they correspond to genet epigen and internal mileu? Kristian: The way I read it, it looks like it corresponds --> Although RT program modifications such as altered exercise volume or intensity may induce more favourable adaptations in some individuals, these benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017].<!-- Add ref for more rt program modifications. Kristian: Added krieger et al. 2009 and schoenfeld et al. 2017 --> For instance, although changes in training volume generally induce greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seem necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? Kristian: Im not sure, are you thinking about "other means than modification of RT" being contradictory of "modifications such as altered exercise volume.." sentence? I think there is a point to establishing that yes, increasing exercise volume is a potent way to increase the RT stimulus in general. However, in cases where this alone fails to elicit true responses, other means are necessary. Or am I missinterpreting this? -->

Nutritional supplements such as protein and creatine effectively optimize RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major supplier of energy to cells via ATP synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] <!--Add ref that indicate high energy requirement for ribo biog possibly (VONWALDEN19;TANAKA18? MOSS. Kristian: These three should do it-->which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by Pol I is considered the rate limiting step in synthesizing new ribosomes [@moss_promotion_1995].  Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target S6K1 serve as a mediator for insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of the translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures[@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

@@

The study was designed as a 12-day double-blinded randomized controlled trial, with an alternating unilateral RT protocol (Figure 1A). Participants were randomly allocated to exercise one leg with a glucose condition and one leg with a placebo condition (Figure 1A).
<!--from the "old" methods: either their dominant or non-dominant leg with glucose or placebo before and after RT, and whether they started RT with glucose or placebo. Thus, participants exercised every day, alternating between e.g., one leg + glucose and the other leg + placebo - this might be TMI, as it's presented in the figure as well?--> One person was exclusively responsible for the randomization code and supplement distribution, blinding both investigators and participants regarding which leg exercised with glucose/placebo. Glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) and placebo (Steviosa, Soma Nordic AS, Oslo, Norway) were masked by mixing with 300ml Fun Light (Orkla, Oslo, Norway). A blinded taste test revealed that the participants were not able to disclose the contents of the provided boluses (30g glucosum monohydricum vs. ~0.3g Stevia rebaudiana extract). To ensure equal conditions during training sessions and strength testing, participants exercised and tested at the same time of day, +/- 1hr with the same supervisor on pairwise consecutive days (i.e. on days 1-2, 3-4, etc.). To further standardize this, participants also recorded and repeated their daily macronutrient intake (protein, fat, carbohydrate) and total calories on pairwise consecutive days.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose/placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of PRO supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously.  During sessions, participants were free to ingest water ad libitum.

@@

Assessment of unilateral one repetition maximum (1RM) leg press and knee extension was conducted at pre-intervention testing during familiarization (Figure 1, Ddays -7 and -5). The participants performed a general warm-up with 10min cycling on an indoor exercise bicycle. A protocol consisting of 1x10, 1x6 and 1x3 repetitions with a load equivalent to ~50-75% of assumed max repetitions, was used as a specific warm-up before each of the tests. All positions were controlled and recorded at the first 1RM test and repeated for the RT sessions. Maximal leg press strength was defined as the maximal load lifted in a controlled fashion, with a knee angle of 45 degrees (start), within a 2.5kg range. Attempts where participants did not reach 45 degrees during the eccentric phase, were not approved. Maximal knee extension testing followed the same specific warm-up and pause protocol and was defined as maximal load lifted in a controlled fashion, reaching full extension of the knee joint, within a 1.25kg range. Attempts with exaggerated hip movement or beneath full extension were not approved. Two minutes of rest were given during the specific warm-up, and three minutes of rest was given between 1RM attempts.

### Resistance training protocol

@@

To measure blood glucose levels with and without glucose ingestion/training, capillary blood was collected from finger draws on days with biopsy sampling. One capillary blood sample was taken on day 1 (Figure 1A) to serve as a baseline. On days 11 and 12 (Figure 1A), capillary blood samples were collected seven times: I) Immediately prior to protein ingestion (0700hrs) II) 45 minutes after protein ingestion (0745hrs) III) 1.5hrs after protein ingestion (0800hrs, i.e., immediately before GLU/PLAC intake), IV) 2hrs after protein ingestion (0900hrs, i.e., immediately before training), IV) in the middle of RT (~0915hrs), V) immediately after training (~0930hrs), and VI) 2hrs after completion of training (~1130hrs). Capillary blood samples were analysed with in-house equipment (BIOSEN C-Line, EKF diagnostic GmbH, Barleben). Venous blood samples were collected from the antecubital vein, coinciding with the capillary samples except 45min after protein ingestion and in the middle of the RT session, to analyse endocrine variables.

@@

After 5 RT sessions (Figure 1D, Post 5RT), there was a significant difference in strength between glucose and placebo,  where placebo decreased strength by 7% more than glucose (Figure 1D, *p* = `r pdiff.c4`). At the other time points, there were no significant differences in mean change between exercising with glucose and placebo (Figure 1D, Post 2RT: *p* = `r pdiff.c2`, Post 4RT: *p* = `r pdiff.c3`, 30min post 6RT: *p* = `r pdiff.c5`, 2h post 6RT: *p* = `r pdiff.c6`, 23h post 6RT: *p* = `r pdiff.c7`). Both RT with glucose and placebo led to significantly reduced strength post-fifth session compared to baseline, by 11 and 18% respectively (Figure 1D, Post 5RT: *p* = 0.000). Comparisons of the acute data gathered from the post-fifth session until and including 23hrs post-sixth session showed a significant mean increase in strength of 5-9% after RT with glucose and placebo 30min after the sixth RT session (Figure 1D, 30min post 6RT: *p* = `r apc2`) and two hours after the sixth RT session (Figure 1D, 2h post 6RT: *p* = `r apc3`). Twenty-three hours after the last (sixth) RT session, strength was unchanged compared to after the fifth RT session (Figure 1D, 23h post 6RT: *p* = `r apc4`).

@@

saveRDS(total_rna_stats, "./data/data-gen/rna/total_rna_stats.RDS")

@@

saveRDS(stat_rrna, "./data/data-gen/rna/stat_rrna.RDS")

@@

stat_west <- lapply(prot_models$prot_models, west_estimate)

saveRDS(stat_west, "./data/data-gen/protein/stat_west.RDS")

p_west <- unlist(lapply(prot_models$prot_models, west_estimate, p = TRUE))

saveRDS(p_west, "./data/data-gen/protein/p_west.RDS")

@@

corr_incr <- round(100 * (exp(coef(summary(mod))[2,1])-1),1)
saveRDS(corr_incr, "./data/data-gen/protein/corr_incr.RDS")

corrp <- sprintf("%.3f", coef(summary(mod))[2,5])
saveRDS(corrp, "./data/data-gen/protein/corrp.RDS")

@@

corrcoef <- coef(summary(corr.mod))
corrp <- sprintf("%.4f", corrcoef[2,5])

@@

The main findings of the present study were that heavy resistance training with glucose did not affect markers of ribosome biogenesis, measured as total RNA, rRNA, and rDNA transcription initiation-associated proteins, compared to RT with placebo. Similarly, RT with glucose in general did not affect markers of skeletal muscle functionality such as strength and recovery, or total training session volume. At one time point, RT with glucose led to less reduction in strength compared to placebo. Levels of plasma glucose and c-peptide were significantly higher before, during and after the RT sessions with glucose. Importantly, there were no differences in daily macronutrient intake on pairwise consecutive days. These findings suggest that the effects of high vs. low glucose/glucose starvation condition on rDNA transcription initiation observed in previous *in vitro* studies [@mariappan_ribosomal_2011; @tanaka_mild_2015] are not translatable to acute effects in human skeletal muscle *in vivo*. Notably, the present study was specifically designed to investigate differences in pre-post changes within participants, and is as such also quite limited to this perspective. Hence, the observed effects of time serve as supplementary to the efficiency of the design, without a negative control group to compare with.

To our knowledge, no previous study has investigated the effects of glucose compared to placebo ingestion with RT on ribosome biogenesis in human skeletal muscle previously, making our hypotheses exploratory and comparisons few. *In vitro* studies on mice glomerular epithelial cells [@mariappan_ribosomal_2011] and human breast cancer cells [@tanaka_mild_2015] have observed glucose-induced UBF-dependent augmentation of ribosome biogenesis [@mariappan_ribosomal_2011], and reduced rRNA transcription by glucose starvation [@tanaka_mild_2015]. These findings do imply an important role for glucose in the initiation of rDNA transcription, the rate-limiting step of ribosome biogenesis [@mariappan_ribosomal_2011; @tanaka_mild_2015]. In addition, Hillier et al. found physiological hyperinsulinemia to stimulate p70S6K phosphorylation in human skeletal muscle [@hillier_physiological_2000]. Moreover, robust increases in markers of ribosome biogenesis such as 47S pre-rRNA and mature rRNA can be expected after an acute bout of RT [@figueiredo_impact_2016] or a short period of RT [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022]. Therefore, we hypothesized that if glucose can, directly or indirectly via insulin, stimulate the initiation of rDNA transcription, five sessions of RT with glucose compared to placebo will elicit this effect. Further, glucose ingestion increases energy levels, leading us to hypothesize that high glucose vs. placebo condition would stimulate energy-sensitive pathways such as PIH1, mTORC1, ERK1/2, AMPK and SIRT1 [@kim_nutrient_2013; @mariappan_ribosomal_2011; @tanaka_mild_2015; @zhai_human_2012]. Despite previously reported upregulation in PIC assembly due to high-glucose mediated mTORC1, ERK1/2 and PIH1 or low-glucose mediated AMPK and SIRT1 activation  [@hoppe_amp-activated_2009; @kim_nutrient_2013; @mariappan_ribosomal_2011; @tanaka_mild_2015; @zhai_human_2012], the present study showed no signs of such effects of glucose vs. placebo conditions. Resistance training, irrespective of condition, yielded a robust accumulation of total RNA and expression of rRNA, in line with previous observations [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022]. Due to the exploratory nature of our hypotheses, this was hardly a very surprising result. Indeed, previous studies have used cell cultures from yeast [@zhai_human_2012], rodent [@hoppe_amp-activated_2009; @mariappan_ribosomal_2011] or human cells [@tanaka_mild_2015] and are, as such, not directly comparable to human skeletal muscle cells. Importantly, previous studies investigated high vs. low glucose conditions [@mariappan_ribosomal_2011], or high glucose vs. glucose starvation [@hoppe_amp-activated_2009; @tanaka_mild_2015], while the present study aimed to compare high glucose condition to placebo, with a matched daily macronutrient and energy intake. Therefore, the comparison made in the present study was high plasma glucose levels vs. normal plasma glucose levels, to investigate the effect of glucose *per se* and not intracellular energy status. Thus, while glucose ingestion presumably is important in supplying energy for growth-inducing processes such as ribosome biogenesis [@figueiredo_regulation_2019; @kusnadi_regulation_2015; @sebata_control_2018] there is no apparent effect of ingesting glucose *per se* on markers of ribosome biogenesis.
